### 2018-08-03 pipeline.tex
Further proofreading

A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016}, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) protein inference, (IV) \ac{MBR} and feature extraction, and (V) relative quantification. Its output can be passed to follow-up analyses in online databases or Bioconductor to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results were comparable to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

@@

The proteome benchmark dataset from \cite{Cox2014} was reanalyzed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{Link to the PRIDE repository entry https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. Briefly, the \textit{Homo sapiens} and \textit{E. coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination. Moreover, each of the 3 replicates of the 2 conditions was analyzed over 24 fractions. This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files. One file was missing in the repository. The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{ThermoRawFileParser repository https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format. These files were used as input for the Compomics+Rob pipeline.

The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{Link to the supplemental data http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file. The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.

@@

The moFF command line utility \cite{Argentini2016} was applied to (I) perform match-between-runs and (II) extract MS1 apex intensity of each peak cluster. This required passing the original RAW files, together with the Default PSM report from PeptideShaker. Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.

@@

Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file. Intensities were $\log_2$ transformed and averaged. The difference between conditions H and L was taken as estimate of the log2FC. Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R. P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.

@@

The pipeline consists of a series of bash, Python and R scripts calling the tools published by the Computational Omics and Systems Biology group at the University of Ghent (see figure \ref{fig:pipeline}).

@@

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities carried out the required search results validation. As expected, matches to the target and decoy databases exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation} A). Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1\% FDR filter. The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{mbr_combined}
\caption[Proteome benchmark MBR results]{Match Between Runs with moFF. \textbf{A} Count of identifications on each run segregated by source. More than 4k spectra were identified and validated by SearchGUI+peptideShaker. In this particular case, hundreds of new identifications were accomplished. \textbf{B} Match visualization. Every dot represents a peptide shared across 2 runs. The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis. The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
\label{fig:mbr}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{curves_plot}
\caption[Classifier evaluation: ROC and PR curves]{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available. \textbf{Comp+Rob} is the pipeline presented in the previous sections. \textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine. Finally, \textbf{MQ+Rob} blends MaxQuant upstream analysis with the MSqRob quantification engine. \textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively. \textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
\label{fig:roc_curves}
\end{figure}

Some of the most frequent evaluations of the performance of a binary classifier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable. Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.

@@

The result of this analysis is shown in figure \ref{fig:roc_curves}. Remarkably, Comp+Rob achieved similar results to those resulting from MQ+LFQ. Nonetheless, the latter performed clearly better when the \ac{log2FC} filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant. In all cases MQ+Rob was as good as MQ+LFQ, if not much better.

@@

In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A). Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins. The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detected by MQ+Rob too. Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power. Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected. Even though the observed correlation was low, the estimated log2FC was always greater than 1. The MQ+Rob pipeline systematically estimated \ac{log2FC} values greater than the other 2.

@@

The low attained matching rate manifests existing room for improvement in the currently available tools. As explained in section \ref{sec:search_engines}, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats. Only MS-GF+ was used in this work for simplicity. Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}. Their usage is predicted to improve identification rates and quality of the matches.

The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM. New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}. Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.

@@

MSqRob treats fractions as a random effect, as explained in section \ref{subsec:quantification}. As a consequence, the statistical model in MSqRob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit{E. coli} proteins. This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis. While no \textit{E. coli} proteins fell in this category in the MQ+MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.

@@

The SearchGUI and PeptideShaker recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{Database formatting guide https://github.com/compomics/searchgui/wiki/DatabaseHelp}}. This is due to the software making use of the annotations to improve the protein inference step. In any case, a consistent formatting in the FASTA databases is required to used the tools. Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with these and many other tools.

>>>>

A pipeline making use of the tools SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016}, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) protein inference, (IV) \ac{MBR} and feature extraction, and (V) relative quantification. Its output can be passed to follow-up analyses in online databases or Bioconductor to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results were comparable to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

@@

The proteome benchmark dataset \cite{Cox2014} was reanalyzed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{Link to the PRIDE repository entry https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. Briefly, the \textit{Homo sapiens} and \textit{E. coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination. Moreover, each of the 3 replicates of the 2 conditions was analyzed over 24 fractions. This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files. One file was missing in the repository. The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{ThermoRawFileParser repository https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format. These files were used as input for the Compomics+Rob pipeline.

The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{Link to the supplemental data http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file. The results of the MQ+Rob pipeline employed the Levenberg-Marquardt-minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.

@@

The moFF command line utility \cite{Argentini2016} was applied to (I) perform match-between-runs and (II) extract MS1 apex intensity of each peak cluster. This required passing the original RAW files, together with the Default PSM report from PeptideShaker. Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity for each sample in a column

@@

Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file. Intensities were $\log_2$ transformed and averaged. The difference between conditions H and L was taken as estimate of the log2FC. Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R. P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R with the "fdr" method.

@@

The pipeline consists of a series of bash, Python and R scripts calling the tools published by the Compomics and StatOmics groups at the University of Ghent (see figure \ref{fig:pipeline}). A parameter file must be created before running the pipeline.

@@

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities carried out the required search results validation. As expected, matches to the target and decoy databases exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation} A). Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1\% FDR filter (see figure \ref{figure:qc_validation} C). The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{mbr_combined}
\caption[Proteome benchmark MBR results]{Match Between Runs with moFF. \textbf{A} Count of identifications on each run segregated by source. More than 4k spectra were identified and validated by SearchGUI+peptideShaker. In this particular case, hundreds of new identifications were accomplished via moFF. \textbf{B} Match visualization. Every dot represents a peptide shared across 2 runs. The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis. The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
\label{fig:mbr}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{curves_plot}
\caption[Classifier evaluation: ROC and PR curves]{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available. \textbf{Comp+Rob} is the pipeline presented in the previous sections. \textbf{MQ+LFQ} consists of a standard MaxQuant upstream analysis combined with its LFQ quantification engine. Finally, \textbf{MQ+Rob} blends MaxQuant upstream analysis with the MSqRob quantification engine. \textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively. \textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
\label{fig:roc_curves}
\end{figure}

Some of the most frequent evaluations of the performance of a binary classifier are the Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable. Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.

@@

The result of this analysis is shown in figure \ref{fig:roc_curves}. Remarkably, Comp+Rob achieved similar results to those resulting from MQ+LFQ. Nonetheless, the latter performed clearly better when the \ac{log2FC} filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant. In all cases MQ+Rob was the best performing classifier.

@@

In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A). Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins. The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detected by MQ+Rob too. Remarkably, 517 proteins were detected by MQ+Rob alone, showcasing its resolving power. Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected. Even though the observed correlation was low, the estimated log2FC was always greater than 1. The MQ+Rob pipeline systematically estimated \ac{log2FC} values greater than the other 2.

@@

The low attained matching rate manifests existing room for improvement in the currently available tools. As explained in section \ref{sec:search_engines}, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats. Only MS-GF+ was used in this work for simplicity. Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{Issue on how to add IdentiPy\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}. Their usage is predicted to improve identification rates and quality of the matches.

A key reason on why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM. New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}. Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.

@@

If fraction aggregation is not supported, MSqRob can be set to treat fractions as a random effect, as explained in section \ref{subsec:quantification}. However, if this is done, the statistical model in Comp+Rob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit{E. coli} proteins. This is clearly manifested in figure \ref{fig:combined_plot} B, which shows many bacterial proteins were assigned  a \ac{log2FC} of 0, indicating MSqRob did not find enough evidence to discard the null hypothesis. While no \textit{E. coli} proteins fell in this category in the MQ+MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.

@@

The SearchGUI and PeptideShaker developers recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{Database formatting guide https://github.com/compomics/searchgui/wiki/DatabaseHelp}}. This is due to the software making use of protein annotations to improve the protein inference step. In any case, a consistent formatting in the FASTA databases is required to use the tools. Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with nost just these but many other tools too.

### 2018-08-03 thp1.tex
Further proofreading

\item Propose a function inference analytical pipeline to complete the proteomics workflow.

@@

THP-1 cells were grown in RPMI1640 (Sigma) supplemented with 10\% FBS (SeraLab) and 1\% pen/strep (Gibco). Stimulation of THP-1 cells was done in 48-well plates (Gibco) with 2x106 cells/well. Cells were exposed to 500 ng/mL LPS (E. coli O111:B4) (+LPS) or nothing (-LPS) for 48 hours in 37 degrees, 5\% \ce{{CO}_2}. After stimulation supernatants were collected and cell pellets were washed in PBS (Gibco). Both supernatants and pellets were frozen (-20 degrees). This work was carried out by Esben G. W. Schmidt. Only the pellet samples were considered for the analysis.

%\subsection{Mass Spectrometry analysis}

@@

 deveThe resulting .RAW files were processed with the two different pipelines introduced in the present work. (I) Compomics+MSqRob and (II) Compomics+BQ (Compomics and BayesQuant). The search settings were identical to those used in chapter \ref{chap:pipeline}, except for the proteome databases employed, which consisted of \textit{Homo sapiens} only. The validation filters, \ac{MBR}, MS1 Apex intensity, and MSqRob preprocessing parameters were also set to those used previously, for simplicity. The two pipelines used thus diverged only in the quantification engine, being identical in all previous steps.

>>>>

\item Propose a biological inference pipeline to complete the proteomics workflow.

@@

THP-1 cells were grown in RPMI1640 (Sigma) supplemented with 10\% FBS (SeraLab) and 1\% pen/strep (Gibco). Stimulation of THP-1 cells was done in 48-well plates (Gibco) with $2 \times 10^6$ cells/well. Cells were exposed to 500 ng/mL LPS (E. coli O111:B4) (+LPS) or nothing (-LPS) for 48 hours in 37 degrees, 5\% \ce{{CO}_2}. After stimulation supernatants were collected and cell pellets were washed in PBS (Gibco). Both supernatants and pellets were frozen (-20 degrees). This work was carried out by Esben G. W. Schmidt. Only the pellet samples were considered for the analysis.

\subsection{Mass Spectrometry analysis}

Peptide digest of THP-1 samples were prepared using a filter assisted sample preparation protocol (FASP). Online desalting and nanoscale LC separation of tryptic peptides was performed with an Dionex Ultimate 3000 Dual –column nano-liquid chromatography system (Thermo Scientific) equipped with two Acclaim PepMap nano-trap columns, 20 mm, 75µ I.D., 3µ particles and two analytical Acclaim PepMap C18, 150 mm, 75µ capillary columns with 2µ particles (Thermo Scientific). Mobile phase A was 0.1\% formic acid (FA) in water and mobile phase B was 0.1\% FA in acetonitrile. Peptides were separated using a 60 min gradient from 5\% to 30\% mobile phase B at 350 nl/min and 35°C injecting samples on the two analytical columns in sequential randomized design. Eluting peptides were analyzed using data-dependent acquisition on an LTQ-Orbitrap Velos Pro mass spectrometer (Thermo Scientific). Each MS scan (400-1600 m/z) was acquired at a resolution of 30000 FWHM followed by 10 low resolution LTQ MS/MS scans triggered above an intensity of 10000 using collision induced dissociation CID (normalized collision energy 35). This work was carried out by David Gleerup and Anders M. B. Giessing.

%In short the FASP method is a filter aided in-solution sample preparation technique comprising of protein extraction with strong chaotropic reagents such as urea or Guanidine-HCL, protein precipitation with TCA, and digestion using trypsin under denaturing conditions of reduced and alkylated protein.  Tryptic peptides are extracted from the filter and analyzed by LC/MS.

%Acquired MS data was processed and annotated using Genedata Expressionist. Acquired MS/MS spectra were searched against Homo sapiens reference proteome database taken using an in-house Mascot server ver. 2.5.1 (Matrix Science). Searches were conducted as follows: semi-tryptic specificity allowing up to two missed cleavages, 10 ppm peptide and 0.6 Da fragment ion tolerances, with carboiodimethyl of cysteine as fixed modification as well as variable modifications consisting of oxidation of methionine and deamidation (NQ). The final protein list was compiled when 2 or more unique peptides per protein were identified with a with a false Discovery Rate (FDR) cutoff of 1%, estimated using a target-decoy search strategy.

@@

The resulting .RAW files were processed with the two different pipelines introduced in the present work. (I) Compomics+MSqRob and (II) Compomics+BQ (Compomics and BayesQuant). The search settings were identical to those used in chapter \ref{chap:pipeline}, except for the proteome databases employed, which consisted of \textit{Homo sapiens} only. The validation filters, \ac{MBR}, MS1 Apex intensity, and MSqRob preprocessing parameters were also set to those used previously, for simplicity. The two pipelines used thus diverged only in the quantification engine, being identical in all previous steps.
